`treats`: a modular `R` package for simulating trees and traits.

### 2023-05-31 paper/treats_paper.Rmd
First draft done!

@@

@@

multivariate stuff @adams2019multivarPCM

@@

Dynamic ones will be for bd.params = list(speciation = runif)

@@

# Discussion

## Modularity

The modular structure allows

## `treats` compared to other packages

Table 1: comparison between the `traits` package and other packages.

@@

<!-- Integrating with dispRity -->
<!-- Lineage specific simulations. -->
<!-- Abiotic events -->
<!-- Time depend simulations -->

# Conclusion
The `treats` is modular and nice to use.

Extended manual following recommendations from @cooper2016dark

@@

>>>>

<!-- no more than 3000-4000 words total -->

@@

Because of its modular architecture it is impossible to provide an exhaustive list of all the package functionalities.
The following sections provide an overview of the three main functions displayed in Figure 1 (`make.traits`, `make.modifiers` and `make.events`).

@@

Traits are simulated via the `make.traits` function given one or more processes, number of dimensions per process and some starting value(s).
Essentially, the generation of new trait values is based on a process (a `function`) modifying a trait value (`x0`) relative to some branch length (`edge.length`).
For example, a simple Brownian Motion process could be generated by the function `rnorm` where the it modifies the trait value `x0` relative to the branch length (`edge.length`).
The longer the branch length, the more likely the new trait value will be different from `x0`:

```{r, eval = FALSE}
## Example of a Brownian Motion process
rnorm(n = 1, mean = x0, sd = edge.length)
```

This can be generated in `treats` by providing this process as a function to `make.traits` which automatically verifies the structure of the process by checking the arguments and applicability to the `treats` algorithm:

```{r, eval = FALSE}
## Functionalising the previous example of a Brownian Motion process
my.bm.process <- function(x0, edge.length) {
    rnorm(n = 1, mean = x0, sd = edge.length)
} 
## Creating the traits object
my.trait <- make.traits(process = my.bm.process)
```

Of course biological data can be much more complex and often multivariate [@adams2019multivarPCM] requiring the users to develop more complex function to cater their specific needs.
However, the `treats` package contains a list of pre-built processes that are "ready-to-use":

 * `BM.process` and `OU.process` which are generalised Brownian Motion and Ornstein-Uhlenbeck processes in any number of dimensions (including possible correlation);
 * `multi.peak.process` which is a generalised Ornstein-Uhlenbeck process (uni- or multi-dimensional) which allows for multiple peaks (usually called the "alpha" parameters; @lande1976OU);
 * `repulsion.process` which is a unidimensional process generating trait values that don't overlap with previously generated trait values;
 * `discrete.process` which generates discrete trait values;
 * `no.process` which ignores the branch length (i.e. a time independent process);

This list will be updated through the lifetime of the package and is documented under the generic `?trait.process` function.

Note that the `treats` package primary aim is to generate both a tree and some traits at the same time.
However, it is possible to also just generate traits with a given topology.
This is done through the function `map.traits` that intakes one or more trees (`"phylo"`, or `"multiPhylo"`) and a `"traits"` object generated by `make.traits` as described above.

@@

Modifying the birth-death process can be done in several ways.
Most easily it is done through changing the stopping rules through the `stop.rule` argument (number of total taxa or living ones, or time of the simulation).
Equally straightforward, one can modify the parameters of the birth-death process: the speciation ($\lambda$) and the extinction ($\mu$) ones.
These can be either fixed values (for constant speciation and extinction) or values drawn from distributions.
This can be specified through the `make.bd.params` function.

However, more interestingly, it is possible to directly modify how the birth death algorithm works through `make.modifiers`.
This function allows to directly modify the three main components of the birth-death algorithm as described above.
By default, the algorithm uses the following algorithms:

1. Generating branch length: this is done by drawing a value from an exponential distribution with the rate being function of the current number of lineages scaled by the speciation and extinction parameters

```{r, eval = FALSE}
## The default branch length generation
rexp(1, rate = number_of_lineages * (speciation + extinction))
```

2. Selecting a lineage among the currently living ones: this is done by simply sampling across the available (living) lineages:

```{r, eval = FALSE}
## The default lineage selection
sample(number_of_lineages, 1)
```

3. Choosing whether that lineage goes extinct (becomes a tip) or speciates (becomes a node): this is done by drawing a random number between 0 and 1 and comparing it to the ratio of speciation and turnover (speciation + extinction).
If the random number is smaller than the ratio of speciation and turnover, the lineage speciates, else it goes extinct (not that if extinction is always null, the speciation/turnover is always 1 so that the lineage always speciates as expected):

```{r, eval = FALSE}
## The default speciation/extinction decider
runif(1) <= (speciation / (speciation + extinction))
```

The function `make.modifiers` allows to specifically modify any of these components by providing a different function for each part of the algorithm.
For example, we could modify the the three functions above so that the branch length is not dependent on the number of lineages, the sampling is always the first species available and the extinction is drawn from a normal distribution.
Note that these function need a specific syntax that is detailed in the [`treats` manual](http://tguillerme.github.io/treats.html).
Briefly, the inputs needs to be always as indicated below and the speciation and extinction parameters as described above are actually handled as part of a list (`bd.params$speciation` and `bd.params$extinction`):

```{r, eval = FALSE}
## Lineage independent waiting:
lineage.independent <- function(bd.params,
                                lineage = NULL,
                                trait.values = NULL,
                                modify.fun = NULL) {
    my_rate <- bd.params$speciation + bd.params$extinction
    return(rexp(1, rate = my_rate))
}
## Selecting always the first species
select.first <- function(bd.params,
                         lineage = NULL,
                         trait.values = NULL,
                         modify.fun = NULL) {
    return(as.integer(1))
}
## Random normal speciation
normal.speciation <- function(bd.params,
                              lineage = NULL,
                              trait.values = NULL,
                              modify.fun = NULL) {
    my_turnover <- bd.params$speciation/
                  (bd.params$speciation + bd.params$extinction)
    return(rnorm(1) <= my_turnover)
}

## Creating the modifier object
modified.birth.death <- make.modifiers(branch.length = lineage.independent,
                                       selection     = select.first,
                                       speciation    = normal.speciation)
```

This `modified.birth.death` is a `"modifiers"` object that is automatically checked for syntax and tested and can be used as a `modifiers` argument in `treats`.

@@

The final major argument to be passed to `treats` are the `"events"` objects generated through `make.events`.
This function works similarly than `make.traits` and `make.modifiers` where the user has to describe the desired event using functions.
For this the following information needs to be specified:

 * the `target` designating what the event should be applied to (e.g. `"taxa"` for modifying the number of species, `"traits"` for modifying the traits, etc.)
 * the `condition` which is a function returning a logical value of when to trigger the event (e.g. when reaching a certain number of taxa, after some specific time has ellapsed or if some specific trait value is reached, etc.)
 * the `modification` which is a function that specifically modifies an internal object in the `treats` algorithm.

Because of the modular architecture, is is impossible to provide an exhaustive list of events so you can refer to the [`treats` manual](http://tguillerme.github.io/treats.html) for many different detailed examples.

Briefly though, this is how the mass extinction events are designed in the "brief applied example" above.

```{r, eval = FALSE}
## Creating an extinction that removes species with positive trait values
positive_extinction <- make.events(
    target       = "taxa",
    condition    = time.condition(15),
    modification = random.extinction(0.75))
```

For this event, the target is the number of taxa in the simulations.
This is indicated using the `target = "taxa"` argument.
Then the event is triggered using the argument `time.condition(15)` and modifies the internal `lineage` object using the ``trait.extinction(x = 0, condition = `>=`)`` function.
`time.condition` and `trait.extinction` are both function factories (functions that outputs other functions) which are not gonna be detailed here (see @wickham2019advanced for more details).
Effectively these arguments can be passed directly as standard functions.
For example, to trigger the event when reaching time 15 we can use the following function:

```{r, eval = FALSE}
## Returns TRUE when reaching time 15
reaching.time15 <- function(bd.params, lineage, trait.values, time) {
    return(time > 15)
}
```

This will trigger the `modification` which is equivalent to the following function modifying the `lineage` internal list:

```{r, eval = FALSE}
removing.75.taxa <- function(bd.params, lineage, trait.values) {
    ## Select a portion of the living species to go extinct
    extinct <- sample(lineage$n, round(lineage$n * 0.75))

    ## Update the lineage object
    lineage$livings <- lineage$livings[-extinct]
    lineage$n       <- lineage$n - length(extinct)
    return(lineage)
}
```

Hence, the extinction event described above is equvivalent to the following one:

```{r, eval = FALSE}
## Creating an extinction that removes species with positive trait values
positive_extinction <- make.events(
    target       = "taxa",
    condition    = reaching.time15,
    modification = removing.75.taxa)
```

### Visualising results

The `treats` package also comes with tools to visualise trees and traits together or separately.
This can be done through the generic S3 `plot` function (calling `plot.treats`) and allows to display up to three traits or two traits and time (in 3D) as displayed in Figures 4 and 5.
Note that you can always use the `plot` functions to data not generated by the `treats` function using the `make.treats` function (transforming trees and data into `"treats"` objects).

# Additional information

## Manuals and vignette

As mentioned through this manuscript, the `treats` package comes with internal documentation (e.g. `?treats`) but also with a thorough and extended vignette in a gitbook format: the [`treats` manual](http://tguillerme.github.io/treats.html).
This manual is designed so that it can be regularly updated and enhanced through the lifetime of the package facilitating the interface between methods development and usage [@cooper2016dark].

@@

This paper describes the first version of the `treats` package.
However, I intend, either personally or through collaborations to continuously develop this package.
For example future planned versions will include: lineage specific simulations; abiotic events; time dependent simulations and; a better integration with the `dispRity` package.
This will be done while keeping track change (through `NEWS` file), continuous integration and unit testing. 
Furthermore, I am planning to develop the GitHub page associated with the package to encourage users to share their simulation templates and build library from them to facilitate the work for future users.

## Repeatability and reproducibility
This paper is entirely reproducible from an Rmarkdown document available on [GitHub](https://github.com/TGuillerme/treats/paper/).
The data used for the example above [@beck2014ancient] is available from the `dispRity` package [@dispRity].

# Conclusion
The `treats` package modular architecture allows workers to develop their own specific biological simulation scenarios based on their own specific research question.
The pipeline of the package through the different `"treats"` objects (`"traits"`, `"modifiers"` and `"events"`) also allows workers to generate publication standard results through plotting but also with easily reproducible and reusable scripts.

@@

<!-- 
## `treats` compared to other packages

Table 1: comparison between the `traits` package and other packages. -->
